Hire Ed: Job Market Dynamics for Tenure-Track Faculty Positions in Archaeology

### 2025-01-08 analysis/paper/article.qmd

The aim of this paper is to TEACH JESS HOW TO USE QUARTO AND GITHUB, UNFORTUNATELY FOR JESS. <!-- It would be nice if you could make comments in the visual mode and not just in source mode, unfortunate --> We had two aims: to determine if disciplinary trends can be discerned in the topical, geographic, and method foci of the positions advertised; and to investigate how the requirements for applicants have changed over time. <!-- yes we can make and view comments in visual mode, my mistake to think it was not possible -->

>>>>

<!--Ben, please note that for now I am keeping track of my references. I have bolded them in text below; I think I will want to learn how to use Zotero and incorporate the citations using your arcane Lovecraftian ritual language (BibTex?), so that for our next paper I am capable of participating in the entire workflow with less supervision-->

Attend any workshop, conference, or panel that includes early career researchers and the conversation will be steered inexorably towards the academic job market: who is hiring, who has attained a tenure-track position, and who is out of luck this season. This obsessive focus on career trajectories is warranted, particularly within the discipline of anthropology. Between 1985 and 2014, the number of PhDs produced by American anthropology departments increased by 70% (**Speakman et al., 2018a**), but the number of tenure-track positions advertised has stagnated or declined*.* As of2019, 53.5% of colleges and universities had replaced tenure-eligible lines with contingent positions (**AAUP 2022**); today, 71% of faculty in the US are non-tenure-track(**Culver and Kezar 2022***).* The erosion of permanent jobs in American higher education is linked to a complex intersection of political and economic factors (**Beck, under review**), including decreasing federal support for higher education, concomitant increases in student debt (**Gusterson 2017**), a pronounced shift in university investment from faculty to administration (Graeber 2018: 162–163), and the growing privatization and market orientation of scientific research (**Mirowski 2011**). Whatever the cause, research within anthropology has consistently shown that at most 20% of PhDs in the discipline are obtaining jobs in American anthropology departments (**Speakman et al. 2018b**), while the top-ranking programmes are placing fewer than one in three graduates in tenure-track jobs (**Mackie and Rockwell 2023**).

These cultural and economic shifts are also well encapsulated by anecdata; any conversation with a junior scholar will reveal the impact of these larger professional transformations on personal and professional trajectories. The first author on this paper has applied to five academic positions over the course of his career, while the last author has applied to one hundred. A twenty-fold increase in the number of applications required to obtain a workable long-term academic position may seem preposterous, but these changing requirements for applicants are familiar to anyone who has obtained their PhDs since the Great Recession. This competitive landscape may have deeper roots than the last economic crisis; writing more than three decades ago, Rabinow pinpointed “an awareness… that standards have changed during the last thirty years and the quantitative and qualitative demands for entry into the system are immeasurably higher now” (**1992: 66**).

These higher demands include increased overall productivity, higher mobility, sometimes requiring bothnational and international moves, more time spent in precarious short-term contracts, and the investment of a greater proportion of time and energy in applying for long-term positions. In their survey of early career researchers in European archaeology, for example, Brami et al. found that 52% of respondents who had obtained their PhD at least two years before the survey were on fixed-term contracts, while only 9% were in permanent positions (**2023: Figure 5**). In their words, “the queue of eager post-docs hoping for a long-term appointment is getting longer, as is the average time spent in the ‘queue’” (**2023: 239**). Ribeiro and Giamakis emphasize that this demographic comprises an academic precariat that is essential to the functioning of universities, but consists of a workforce on short-term contracts perpetually “stranded between employment and unemployment” (**2023:10**).

Early career scholars with faculty ambitions are thus stuck in a double-bind, forced to funnel their energy into seemingly endless applications for which they receive little to no feedback and which remain invisible on their CVs. As Dennis et al. underscore, early career researchers are “supplicants”: “There is no way for applicants to point out that the job ads are taking too much time out of the scholarly community’s collective time bank” (**2022: 1**). Indeed, acknowledging the new reality that applying for academic positions is in and of itself a full-time job, a growing number of anthropology graduate programmes are offering professionalization courses that cover the ins and outs of everything from crafting strong cover letters and CVs to performing well in Zoom interviews and teaching demonstrations. These courses often provide valuable training in the hidden curriculum of the academy, but that job applications are so complicated that they require training to navigate suggests more disciplinary attention should be paid to the “market” as a historically contingent set of cultural practices. As Rabinow maintains, “if we want ethical considerations to play a central role in the articulation of truth and power—and I think we do—then bringing such considerations into view is the necessary first step towards recognizing who we are today and setting out on the road to a better place" (**1992: 71**) **.**

We argue that to understand the dynamics of hiring in academic archaeology, we must start at the beginning, with the job ad itself. There is an abundance of recent work on who does and does not get hired into tenure-track positions in anthropology and archaeology (Kawa 2019; Mackie and Rockwell 2023), but less scholarship on how those hires unfold. Research on hiring dynamics has amply demonstrated that *where* you get your PhD has an enormous impact on if and where you get a tenure-track position (Kawa et al. 2019; Mackie and Rockwell 2023), but how about *what* you study? Are there topical, methodological, or regional specialties that are in consistent demand, or is there continual flux generated by intellectual fads, economic incentives, and political attention? Are job applications more demanding now than they were in the past, or is this a specious perception generated by the sheer volume of applications necessary to obtain a long-term academic position?

To answer these questions, this paper explores the demand-side of the academic job market for archaeologists in the United States. Our study had two aims: first, to determine if disciplinary trends could be discerned in the topical, geographic, and methodological foci of the positions advertised over a ten-year period, and second, to investigate how requirements for applicants have changed over time.
